Multiresolution Neural Field Estimation

### 2011-03-01 Multiresolution_framework/ltx/manuscript.tex
edits to intro
edits to discussion

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Multiresolution Neural Field Estimation }
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{In no particular order: \parham{P. Aram}, \mike{M. Dewar}, \dean{D.R. Freestone}, \ken{K.Scerri}, D.B. Grayden and V. Kadirkamanathan % <-this % stops a space
\thanks{P. Aram* and V. Kadirkamanathan are with the Department of Automatic Control and Systems Engineering, University of Sheffield, Sheffield, S1 3JD, U.K. (e-mail:p.aram@sheffield.ac.uk; visakan@sheffield.ac.uk).}% <-this % stops a space
\thanks{M. Dewar is with the School of Informatics, University of Edinburgh, Edinburgh, EH8 9AB, U.K. (e-mail:mike.dewar@inf.ed.ac.uk)}
\thanks{D.\ R.\ Freestone and D.\ B.\ Grayden are with the Department
of Electrical and Electronic Engineering, The University of Melbourne, and The Bionic Ear Institute, VIC, Australia,  {\tt\small dfreestone@bionicear.org, grayden@unimelb.edu.au}}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

@@

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
%  \begin{center} \bfseries EDICS Category: SSP-IDEN \end{center}
%  \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

% We have a clear definition and where to stimulate and how to stimulate.\\
% \cut{This color is used to indicate bits we can get rid of if we don't have enough space, use the command: \textbackslash cut$\left\lbrace  \right\rbrace$}

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, \cut{we denote the current time step by $t$ and the future time step by $t+1$.} The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}.

The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.

\section{MRA of the IDE in State-Space}
The multiresolution approximation (MRA) of the neural IDE is obtained by decomposing both the field, $v_t(.)$, and the connectivity kernel, $w(.)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(\mathbf{r})$ and a mother wavelet $\psi(\mathbf{r})$. Considering a one-dimensional field, the connectivity kernel is decomposed by,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters respectively. The field is decomposed as
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $ x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion at time $t$.

@@

The multiresolution approximation can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{align}
N_{m}\left(r\right)=&\left(N_{m-1}\ast N_{1}\right)\left(r\right)\nonumber \\
=&\int_0^{1} N_{m-1}\left( r-r'\right)dr' \quad m>1,
\label{SplineConvolutionIntegral}
\end{align}
where $\ast$ denotes convolution and $N_1\left(s\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Eq.~\ref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ times convolution of indicator function with itself since
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of Eq.~\eqref{eq:BsplineConvolution} is (see \cite{Chui1992} for proof) 
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. The $m^{th}$ order B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} q_n N_{m}\left(2r-n\right) \\
 q & _n = \frac{\left(-1\right)^n}{2^{m-1}} \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting properties Eqs.~\ref{eq:BsplineConvolution} and \ref{eq:BsplineInnerProduct} the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta} can be computed. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale.  In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta}.

\section{State and Parameter Estimation}
The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ block of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\mike{you need to define and sort out the $[]_{ij}$ notation!}
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.%\cut{The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.} 
The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel (RTS)\cite{RAUCH1965} smoother outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t-1},\mathbf{x}_{t})$ 
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}
Hence, the algorithm has two distinct steps: the E-step which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step which updates the paramter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge.

@@

This work extends previous work by... alternative ARMA model...

@@

% In order to estimate the spatial mixing kernel and the spatio-temporal field, the  kernel and the field are decomposed using
% Note that the number of translation operations depends on the spatial range of the data. It follows that the total number of terms in (\ref{1DKernelDecompositionForSimulation}) and (\ref{1DFieldDecompositionForSimulation}) are 8 and 47 respectively. The estimated kernel in \figurename{\ref{fig:SpatialMixingKernelEstimate} } is obtained for the cubic B-spline, with initial scale 0 and maximum scale 1. The EM algorithm is allowed to run for 30 iterations to avoid any issues regarding early stopping, though typically the change in $\parallel \mathbf{A} \parallel_{F}$ falls below $10^{-4} $ after less than 15 iterations. A total number of 500 estimation experiments are performed, where $v_t$ and $w_t$ are re-generated each trial.

%  The performance measure employed here is the Mean Integrated Squared Error (MISE) defined as
% \begin{equation}
%  \mathbf{E}\left[\int_{\mathcal{S}}\left[e\left(s\right) \right]^2ds  \right]
% \label{MISE} 
% \end{equation}
%  where  $e\left(s\right)$ is the error between the unknown function, $f\left( s\right) $ and its estimate, $\hat{f}\left( s\right) $. In this example the original kernel and the field are known and hence the integral in  (\ref{MISE}) can be computed analytically. For spatial mixing kernel we have
% \begin{eqnarray}
%  \lefteqn{\int_{\mathcal{S}}\left[e\left(s\right)\right]^2ds}\nonumber \\  &=\left(\boldsymbol{\theta}-\hat{\boldsymbol{\theta}}\right)^\top\left[ \int_{\mathcal{S}}\boldsymbol\lambda\left(s\right)\boldsymbol\lambda\left(s\right)^\top ds\right] \left(\boldsymbol{\theta}-\hat{\boldsymbol{\theta}}\right)&
% \label{MISEKernel}
% \end{eqnarray}
%   Here, we are particularly interested in assessing the performance improvement as the scale of the details in MRA is increased. Performance based on (\ref{MISEKernel}) at different spatial scales and over 500 trials have been evaluated. The MISE of the model at $j=0$ is 0.024 and reduces to 0.005 when $\psi_{4;1,k}$ terms are also included in the model, in fact estimation performance is improved by a factor of 4.8. The coarsest approximation of the kernel along with its approximation at $j=1$ are illustrated in \figurename{\ref{fig:SpatialMixingKernelEstimate}. It can be seen that the model is able to estimate both slowly and rapidly varying segments of the  kernel with very high accuracy.  Field estimates  at a selection of time instants are shown in \figurename{\ref{fig:1DFieldEstimation}}. It is clearly observed that the coarse and fine features of the original field have been captured accurately over regions where observations are available. Observation locations are generated randomly, as shown in \figurename{ \ref{fig:FieldVariance}}. Assuming the observed field can be fully described by the  decomposed IDE model, the variance of the field can be computed to express the  uncertainty of the estimation, at each point in space and at time $t$ we have 
% \begin{equation}
%  \text{Var}(\hat z_t(s_i))=\boldsymbol\mu^{\top}(s_i) \mathbf P_{t|T}\boldsymbol\mu(s_i)
% \label{eq:FieldVariance}
% \end{equation}
% where $\mathbf P_{t|T} $ is the covariance matrix associated with the  state estimates obtained from the RTS smoother. Variance based on (\ref{eq:FieldVariance}) at each spatial location and at $t=50$ is shown in \figurename{\ref{fig:FieldVariance}}. The uncertainty of the estimation is higher at almost  every spatial location at $j=1$ compared to $j=0$. This is to be expected as there are more wavelet basis functions at higher resolution and hence more associated weights need to be estimated. Peaks in the variance of the field coincide with regions where no observations have been made. Increasing the number of observation locations, or using equally spaced observation locations would reduce the variance of the field estimation.
% \begin{figure}[!h] 
% \centering
% \includegraphics[width=0.5\textwidth]{./Graph/aram4.eps}
% \caption{Variance of the field at each spatial location for $j=0$ (dotted line) and $j=1$ (solid line). The observation locations are described with a cross.}
% \label{fig:FieldVariance}
% \end{figure} 
% \begin{table}[!h]
% \renewcommand{\arraystretch}{1.3}
% \caption {MISE  at particular scale between measured and estimated 2D-field  at a particular time}
% \label{table:ValidationResultField}
% \centering
% \begin{tabular}{cccc}
% \hline \hline
% \multicolumn{2}{r}{Time}&\multicolumn{2}{c}{MISE} \\
% \cline{3-4}
% &&$j=0$&$j=1$ \\ 
% \hline 
% &$5$ &$5.424$&$0.609$ \\
% &$20$ &$5.422$&$0.596$ \\
% &$50$ &$5.519$&$0.626$ \\
% &$100$ &$5.488$&$0.611$ \\
% &$150$ &$5.494$&$0.587$ \\
% \hline \hline
% \end{tabular}
% \end {table}
% \section{Conclusion}

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Multiresolution Neural Field Estimation }

\author{In no particular order: \parham{P. Aram}, \mike{M. Dewar}, \dean{D.R. Freestone,~\IEEEmembership{Graduate Student Member,~IEEE,}}, \ken{K.Scerri}, D.B. Grayden,~\IEEEmembership{Member,~IEEE,} and V. Kadirkamanathan % <-this % stops a space
\thanks{P. Aram* and V. Kadirkamanathan are with the Department of Automatic Control and Systems Engineering, University of Sheffield, Sheffield, S1 3JD, U.K. (e-mail:p.aram@sheffield.ac.uk; visakan@sheffield.ac.uk).}% <-this % stops a space
\thanks{M. Dewar is with the School of Informatics, University of Edinburgh, Edinburgh, EH8 9AB, U.K. (e-mail:mike.dewar@inf.ed.ac.uk)}
\thanks{D.\ R.\ Freestone and D.\ B.\ Grayden are with the Department
of Electrical and Electronic Engineering, The University of Melbourne, and The Bionic Ear Institute, VIC, Australia,  {\tt\small dfreestone@bionicear.org, grayden@unimelb.edu.au}}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

@@

@@

\IEEEpeerreviewmaketitle

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, \cut{we denote the current time step by $t$ and the future time step by $t+1$.} The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function.

The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.
\section{MRA of the IDE in State-Space}
The multiresolution approximation (MRA) of the neural IDE is obtained by decomposing both the field, $v_t(.)$, and the connectivity kernel, $w(.)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(\mathbf{r})$ and a mother wavelet $\psi(\mathbf{r})$. Considering a one-dimensional field, the connectivity kernel is decomposed by,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters respectively. The field is decomposed as
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $ x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion at time $t$.

@@

The multiresolution approximation can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{align}
N_{m}\left(r\right)=&\left(N_{m-1}\ast N_{1}\right)\left(r\right)\nonumber \\
=&\int_0^{1} N_{m-1}\left( r-r'\right)dr' \quad m>1,
\label{SplineConvolutionIntegral}
\end{align}
where $\ast$ denotes convolution and $N_1\left(s\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Eq.~\ref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ times convolution of indicator function with itself since
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of Eq.~\eqref{eq:BsplineConvolution} is (see \cite{Chui1992} for proof) 
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. The $m^{th}$ order B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} q_n N_{m}\left(2r-n\right) \\
 q & _n = \frac{\left(-1\right)^n}{2^{m-1}} \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting properties Eqs.~\ref{eq:BsplineConvolution} and \ref{eq:BsplineInnerProduct} the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta} can be computed. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale.  In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta}.
\section{State and Parameter Estimation}
The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ block of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\mike{you need to define and sort out the $[]_{ij}$ notation!}
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.%\cut{The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.} 
The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel (RTS)\cite{RAUCH1965} smoother outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t-1},\mathbf{x}_{t})$ 
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}
Hence, the algorithm has two distinct steps: the E-step which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step which updates the paramter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge.

@@

This work extends previous work by... alternative ARMA model... The MRA IDE can be applied to fluid dynamics, weather systems, ....

@@
